Efficiency

### 2018-01-08 DraftExec.tex
I'm at hotel load sensitivity

In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.  The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.

@@

The lower propeller curve shows efficiency still increasing up to around 65\% at 20kt, which might be consistent with a propeller optimised for conventional ship speeds, but probably not likely for a specialised submarine propeller.  For all propeller curves, sharply declining efficiency has been assumed at very low speeds, (2kt or less), in order to reflect the dip seen below 5kt in the efficiency curve seen in Figure \ref{fig:BMTEfficiency.png}, assuming that this is driven by the decline in efficiency shown in Figure \ref{fig:PropDuct.png} at advance ratios approaching 1.  This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.  It is unlikely that unless some other significant factor not discussed in this paper that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt. Hence, the low propeller curve is considered relatively unlikely, but useful for investigating sensitivity to errors or adjustments.

@@

\subsection{Hotel Load Assumptions}
It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.  We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are few.  It is also likely that the actual hotel load varies considerably as different systems are switched on and off, some of which might be discretionary in combat scenarios (e.g. hot water for washing), and others (such as the combat systems) might have increased demand during crucial phases of combat operations.  It is important to note that the combat system that has been selected for Australias FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more. \cite{patrick2011}.  Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.  A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to given an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines. Precise energy density of these batteries is not publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two \parencite{greenfield2016}.

@@

In the online version of the web-app, alternative reference points or known facts relating a given power output to a speed can be used in order to solve for this constant $C_d$.  For instance, one might assume (as done by \cite{davies2017}) that the hotel load might match the propulsive demand at some given speed, for some given hotel load.  However, given that the precise hotel load is generally not known, and highly variable, and there is potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.

Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumptio at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.  It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".  Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt. Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.

@@

The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.  Under central assumptions, the propeller driven submarine can reach over 540nm, at any speed between 4 and 5kt. The pump-jet driven submarine reaches over 360nm between 3 and 6kt. A difference 180nm dived range would presumeably be of considerable tactical significance.

@@

\subsection{Sensitivity of Curve Selection}
As can be seen from the comprehensive comparison in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}, the shape of curves selected has very different levels of impact at different speeds.  Key observations include how small the impact of efficiency variation to eithre curves tends to be at very low speeds (under 2kt).  This is due to hotel load becoming clearly dominant at such speeds.  The relative difference between the lines also diminishes somewhat towards higher speeds.  More striking, however, is the very dramatic reduction in both range and endurance at higher speeds. At a speed of 20kt, a propeller driven sumbarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.  A difference of about half an hour would presumeably be of less tactical significance. At any speed above 10kt, neither submarine would be able to remain dived for much more than 24 hours (25.4 hours for the propeller system, 18.3 for the jet).

@@

A further feature of the results is that the relative difference in performance is much more sensitive to changes in the efficiency of the jet at the relevant speeds, owing to their larger proportional impact at low efficiencies.  Whilst a smoothly declining curve from around or over 50\% at 20kt to zero at 0kt, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.  This can be seen in an overlaid plot of the proportional differences in total power demand (hence endurance and range) for the different pairs of curves as in Figure \ref{fig:all_proportions}.

@@

The proportional change in overall power requirement (directly determining both range and endurance) shows that our central scenario shows a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt. It is interesting to note that the proportional change is greatest slightly above the speed where maximum range is achieved, and the point of peak advantage moves steadily towards higher speeds as the difference in efficiency between the systems increases.

As can be seen, even compred to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.  It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.  Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology.

@@

It's important to acknowledge that such substantial differences in hotel load merit serious consideration, both in terms system design, as well as for variance in operation of a submarine with a given system design.  As has been discussed, the precise hotel load of submarines is not generally publically available, which justifies senitivity testing around a relatively wide potential range.  The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking. In addition, it broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.

Furthermore, the hotel load may be varied considerably at the captain's discretion depending on operational requirements.  Not all sensors or components of the combat system need to run at all times, for example during long transits in order to reach an operational area, or during a long dive in order to escape an operational area once a large strike has already been conducted.  In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.  Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) achievable.  The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.

>>>>

In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.  The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.  The technologies naturally will not be identical.  It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed flow pump, but this difference in turbomachinery is unlikely to result in a overall loss of efficiency \parencite{hamilton1997}.  The intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.  It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine. However, being  for a warship designed to travel at similar speeds to a nuclear submarine, with relatively unconstrained propeller size, and the pumpjet ejecting into the water, the technologies are likely to be far closer than any other publically available comparison. Overall, however, its considered that having an experimtally verified curve comparing very similar technologies as a direct function of waterspeed provides a far more certain basis for comparison than attempting to derive different curves \textit{ab initio} using theortical dimesnionless curves attempting to estimate how the proposed designs might correspond to them.

@@

The lower propeller curve shows efficiency still increasing up to around 65\% at 20kt, which might be consistent with a propeller optimised for conventional ship speeds, but probably not likely for a specialised submarine propeller.  For all propeller curves, sharply declining efficiency has been assumed at very low speeds, (2kt or less), in order to reflect the dip seen below 5kt in the efficiency curve seen in Figure \ref{fig:BMTEfficiency.png}, assuming that this is driven by the decline in efficiency shown in Figure \ref{fig:PropDuct.png} at advance ratios approaching 1.  This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.  It is unlikely that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt, unless some other significant factor not discussed in this paper affects propellers in these regimes. At very low speeds, some small but sudden changes in efficiency may be possible in either direction, particularly as complex effect with multiple overlaid effects take place, for example increasing laminar flow over the propeller blades leading to lower frictional losses, but potentially greater separation.  However, on balance, the low propeller curve is considered relatively unlikely, but useful for investigating sensitivity to errors or adjustments.

\begin{figure}
\includegraphics[width=\textwidth]{LaminarProp.png}
\caption{At very low speeds changes in regimes such as an increase in laminar flow over the blade section could lead to small changes in efficiency, perhaps explaining the distinct rise at low speeds for the properller at ~5kt in Figure \ref{fig:BMTEfficiency.png}}
\label{fig:LaminarProp.png}
\end{figure}

@@

\subsection{Hotel Load Assumptions}
It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.  We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are scarce.  It is also likely that the actual hotel load varies considerably as different systems are switched on and off, some of which might be discretionary in combat scenarios (e.g. hot water for washing), and others (such as the combat systems) might have increased demand during crucial phases of combat operations.  It is important to note that the combat system that has been selected for Australias FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more. \cite{patrick2011}.  Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.  A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to give an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines. Precise energy density of these batteries is not publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.

@@

In the online version of the web-app, alternative reference points or known facts relating a given power output to a speed can be used in order to solve for this constant $C_d$.  For instance, one might assume (as done by \cite{davies2017}) that the hotel load might match the propulsive demand at some given speed, for some given hotel load.  However, given that the precise hotel load is generally not known and highly variable, and the potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.

Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumption at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.  It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".  Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt. Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.

@@

The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.  Under central assumptions, the propeller driven submarine can reach over 540nm at any speed between 4 and 5kt. The pump-jet driven submarine reaches over 360nm between 3 and 6kt. A difference of 180nm dived range would presumeably be of considerable tactical significance.

@@

\subsection{Sensitivity of Curve Selection}
As can be seen from the comprehensive comparison in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}, the shape of curves selected has very different levels of impact at different speeds.  Key observations include how small the impact of efficiency variation to either curves tends to be at very low speeds (under 2kt).  This is due to hotel load becoming clearly dominant at such speeds.  The relative difference between the lines also diminishes somewhat towards higher speeds.  More striking, however, is the very dramatic reduction in both range and endurance at higher speeds. At a speed of 20kt, a propeller driven sumbarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.  A difference of about half an hour would presumeably be of less tactical significance. At any speed above 10kt, neither submarine would be able to remain dived for much more than 24 hours (25.4 hours for the propeller system, 18.3 for the jet).

@@

A further feature of the results is that the relative difference in performance is much more sensitive to changes in the efficiency of the jet at the relevant speeds, owing to their larger proportional impact at low efficiencies.  Whilst a smoothly declining efficiency curve from around or over 50\% at 20kt to zero at 0kt might seem moderately well defined, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.  This can be seen in an overlaid plot of the proportional differences in total power demand (hence endurance and range) for the different pairs of curves as in Figure \ref{fig:all_proportions}.

@@

The proportional change in overall power requirement (directly determining both range and endurance) shows that in the central scenario a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt. It is interesting to note that the proportional change is greatest slightly above the speed where maximum range is achieved, and the point of peak advantage moves steadily towards higher speeds as the difference in efficiency between the systems increases.

As can be seen, even compared to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.  It is interesting to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.  Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology, as will be discussed later.

@@

It's important to acknowledge that such substantial differences in hotel load merit serious consideration, both in terms system design, as well as for variance in operation of a submarine with a given system design.  As has been discussed, the precise hotel load of submarines is not generally publically available, which justifies senitivity testing around a relatively wide potential range.  The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking. In addition, it is broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.

Furthermore, the hotel load may be varied considerably at the captain's discretion depending on operational requirements.  Not all sensors or components of the combat system need to run at all times, for example during long transits in order to reach an operational area, or during a long dive in order to escape an operational area once a large strike has already been conducted.  In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.  Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) is achievable.  The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.
